---
title: "data clean"
author: "Huiyi Zhu"
date: "2025-12-04"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## prepare

```{r}
library(tidyverse)
library(janitor)
library(ggplot2)
library(patchwork)
```
outcome: attendance 
variables: season, event type, borough, location type, time of date



Miranda Chen: model (before 11:59PM Friday)
Huiyi Zhu: clean data (date:2022.5-present, morning5-12 afternoon 12-6 night 6-1 midnight?) + basic descriptive analysis 
Yiyang Jiao: report+video (9:00PM Sat)
Xiaoqian Lyu: report
Yuyue Ma: web(sat) 


## Data Cleaning
```{r}
parkevent = read.csv("Parks_Special_Events_20251204.csv") |>
  janitor::clean_names() |>
  mutate(
    datetime = mdy_hms(date_and_time),
    date     = as_date(datetime), 
    year  = year(datetime),
    month = month(datetime, label = TRUE, abbr = TRUE),
    day = day(datetime),
    time = format(datetime, "%H:%M:%S"),
    season = case_when(
      month %in% c("Mar","Apr","May") ~ "Spring",
      month %in% c("Jun","Jul","Aug") ~ "Summer",
      month %in% c("Sep","Oct","Nov") ~ "Fall",
      month %in% c("Dec","Jan","Feb") ~ "Winter"), 
      season = factor(season,
        levels = c("Spring", "Summer", "Fall", "Winter")),
    hour = as.numeric(format(datetime, "%H")),
    time_period = case_when(
      hour < 5 ~ "Midnight",
      hour >= 5  & hour < 12 ~ "Morning",
      hour >= 12 & hour < 18 ~ "Afternoon",
      hour >= 18 & hour < 23 ~ "Night",
      hour >= 23   ~ "Midnight"),
   time_period = factor(time_period,
                     levels = c("Morning", "Afternoon", "Night", "Midnight")),
    attendance = as.numeric(attendance))|>  
  mutate(
    aud_list = str_split(audience, ";|,"),
    aud_list = map(aud_list, ~ str_trim(.x))) |>
  mutate(
    has_children = map_lgl(aud_list, ~ any(.x %in% c("Children", "Kids", "Tot"))),
    has_seniors  = map_lgl(aud_list, ~ any(.x %in% c("Seniors", "Senior"))),
    has_genpub   = map_lgl(aud_list, ~ any(.x == "General Public")),
    
    kids_friendly   = has_children | has_genpub,
    senior_friendly = has_seniors  | has_genpub,
    adults_only     = !kids_friendly) |>
  select(
    -datetime,
    -unit,
    -group_name_partner,
    -date_and_time,
    -classification,
    -source,
    -hour,
    -event_name,
    -location) |> 
  select(
    date, year, month, day, time, season,
    everything()) |> 
    filter(date >= as.Date("2021-06-01"),
           event_type != "",
           category != "",
           attendance != "")
```

We cleaned the dataset to prepare it for analyzing attendance as the primary outcome and key predictors including season, event_type, borough, location_type, and time_period. Date-time information was parsed into date, year, month, and day to support temporal analyses, and time_period was created to evaluate differences across times of day. We filtered observations to the NYC fully reopened period after COVID (date >= "2021-06-01") to avoid pandemic-related disruptions. Columns not relevant to the analysis were removed, and rows with empty event_type, category and attendance values were dropped. After cleaning, we have `r nrow(parkevent)` observations and `r ncol(parkevent)` variables. These steps produced a tidy dataset aligned with our study.

How we cleaned the audience type: We first standardized the audience field by splitting each entry into individual audience categories (e.g., “Children,” “Adults,” “Seniors,” “General Public”). Based on these cleaned components, we created three binary indicators to capture the accessibility of each event to key demographic groups.

An event was classified as kids-friendly if its audience included Children/Tot or was labeled as General Public. Similarly, an event was marked as senior-friendly if Seniors were listed or if the audience was General Public. Finally, we defined adults-only events as those not classified as kids-friendly, meaning no child-oriented groups were present in the audience description.

```{r}
parkevent %>% count(kids_friendly)
parkevent %>% count(senior_friendly)
parkevent %>% count(adults_only)
```




## Descriptive Analysis 
```{r}
parkevent |>
  summarize(
    mean_attendance = mean(attendance),
    median_attendance = median(attendance),
    sd_attendance = sd(attendance),
    min_attendance = min(attendance),
    max_attendance = max(attendance)) |>
  knitr::kable(digits = 2)
```

This table summarizes the distribution of attendance across all post-COVID events, including the mean, median, variability, and range of turnout. These descriptive statistics provide an essential baseline for understanding overall engagement levels and for comparing attendance across seasons, event types, boroughs, and time-of-day categories in subsequent analyses.

```{r}
plot_season =
  ggplot(parkevent, aes(x = season, y = attendance, fill=season)) +
  geom_violin(trim = FALSE, alpha = 0.5) +
  geom_boxplot(width=0.15) +
  coord_cartesian(ylim = c(0,250)) +
  scale_x_discrete(labels = c(
    "Spring" = "Spr",
    "Summer" = "Sum",
    "Fall"   = "Fal",
    "Winter" = "Win")) +
  labs(title = "By Season", x = "Season", y = "Attendance") +
  guides(fill = "none") +
  theme_minimal()

plot_time =
  ggplot(parkevent, aes(x = time_period, y = attendance, fill=time_period)) +
  geom_violin(trim = FALSE, alpha = 0.5) +
  geom_boxplot(width=0.15) +
  coord_cartesian(ylim = c(0,250)) +
  scale_x_discrete(labels = c(
    "Morning"   = "Morn",
    "Afternoon" = "Aft",
    "Night"     = "Ngt",
    "Midnight"  = "Mid")) +
  labs(title = "By Time of Day", x = "Time of Day", y = "Attendance") +
  guides(fill = "none") +
  theme_minimal()

plot_type =
  ggplot(parkevent, aes(x = event_type, y = attendance, fill = event_type)) +
  geom_violin(trim = FALSE, alpha = 0.5) +
  geom_boxplot(width = 0.15) +
  coord_cartesian(ylim = c(0,250)) +
  scale_x_discrete(labels = c(
    "Agency Produced Event" = "Agency",
    "Community Based Event" = "Community",
    "Local Event" = "Local",
    "Open House" = "Open",
    "Tournament/Competition" = "Compet")) +
   guides(fill = "none") +
  labs(
    title = "By Event Type",
    x = "Event Type",
    y = "Attendance") +
  theme_minimal()

plot_borough =
  ggplot(parkevent, aes(x = borough, y = attendance, fill=borough)) +
  geom_violin(trim = FALSE, alpha = 0.5) +
  geom_boxplot(width=0.15) +
  coord_cartesian(ylim = c(0,250)) +
  scale_x_discrete(labels = c(
    "Bronx"        = "BX",
    "Brooklyn"     = "BKN",
    "Manhattan"    = "MAN",
    "Queens"       = "QNS",
    "Staten Island"= "SI")) +
  labs(title = "By Borough", x = "Borough", y = "Attendance") +
  guides(fill = "none") +
  theme_minimal()

plot_location =
  ggplot(parkevent, aes(x = location_type, y = attendance, fill = location_type)) +
  geom_violin(trim = FALSE, alpha = 0.5) +
  geom_boxplot(width = 0.15) +
  coord_cartesian(ylim = c(0,250)) +
  scale_x_discrete(labels = c(
    "Nature Center" = "Nature",
    "Recreation Center" = "RecCen",
    "Playground" = "Playground",
    "Pool" = "Pool",
    "Park" = "Park",
    "Other" = "Other",
    "School" = "School")) +
  labs(
    title = "By Location Type",
    x = "Location Type",
    y = "Attendance") +
   guides(fill = "none") +
  theme_minimal()

patterns_plot =
  ((plot_season |plot_time| plot_borough) /
    (plot_type| plot_location))+
  plot_annotation(
    title = "Attendance Patterns Across NYC Parks Events",
    theme = theme())
patterns_plot

```

Figure displays violin–boxplots of attendance by season, time of day, borough, event type, and location type. These plots summarize both the distribution (violin shape) and central tendency (box plot) for each category, allowing visual comparison of how attendance varies across park events. We restricted the y-axis to 0–250 attendees because most events fall within this range, while a small number of very large events have much higher attendance. Without this limit, those extreme values compress the main body of the data and make differences between typical events difficult to see.


```{r}
plot_time_season =
  parkevent |>
  group_by(time_period, season) |>
  summarize(mean_attendance = mean(attendance)) |>
  ggplot(aes(x = time_period, 
             y = mean_attendance, 
             color = season, 
             group = season)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Mean Attendance by Time of Day Across Seasons",
    x = "Time of Day",
    y = "Mean Attendance",
    color = "Season") +
  theme_minimal()

plot_borough_season =
  parkevent |>
  group_by(borough, season) |>
  summarize(mean_attendance = mean(attendance)) |>
  ggplot(aes(x = season, 
             y = mean_attendance, 
             color = borough, 
             group = borough)) +
  geom_line() +
  geom_point() +
   scale_color_brewer(palette = "Dark2") +  
  labs(
    title = "Mean Attendance by Seasons Across Borough",
    x = "Season",
    y = "Mean Attendance",
    color = "borough") +
  theme_minimal()

Seasonal_plot =
  (plot_time_season / plot_borough_season) +
  plot_annotation(
    title = "Seasonal Attendance Patterns Across NYC Park Events")

Seasonal_plot
```

The first plot shows how mean attendance varies by time of day across seasons. Spring consistently has the highest turnout, especially in the afternoon, while Winter shows the sharpest decline at midnight. Fall attendance remains relatively stable across all times, whereas Summer gradually increases into the evening. The second plot compares mean attendance by seasons across borough. Staten Island and Queens show the highest seasonal peaks, particularly in Spring. Brooklyn and Manhattan maintain moderate attendance across seasons, while the Bronx shows the lowest and least variable averages. Together, these trends highlight clear seasonal and geographic differences in NYC park event engagement.


```{r}

plot_calendar =
  ggplot(parkevent, 
         aes(x = month(date, label = TRUE, abbr = TRUE), 
             y = attendance)) +
  geom_point(alpha = 0.2, color = "gray70") +
  stat_summary(fun = mean, geom = "line", aes(group = 1),
               color = "cornflowerblue", size = 1) +
  stat_summary(fun = mean, geom = "point",
               color = "cornflowerblue", size = 2) +
  labs(
    title = "Attendance Over the Calendar Year",
    x = "Month",
    y = "Attendance") +
  coord_cartesian(ylim = c(0, 250)) + 
  theme_minimal()
plot_calendar
```

This plot shows how park event attendance varies throughout the calendar year by combining individual event points with a smoothed monthly trend line. Grey points display the full distribution of attendance for all events, allowing visualization of variability within each month. Monthly averages are overlaid as a blue line to highlight broader seasonal patterns that daily fluctuations may obscure. The results show a dip in February, rising attendance during spring, a slight decline in midsummer, and another increase in October and December. Using monthly means helps reveal consistent seasonal trends while still preserving the underlying data distribution for context and interpretation.

